Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow

### 2024-03-02 manuscript/Manuscript.qmd
Some typos

---
title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

# Survey coverage and other colour overlayed the municipalities
muniPlot <- tm_shape(muni3) +
  tm_borders() +
  tm_facets(
    by = "Municipality",
    ncol = 3
  ) +
  tm_shape(terrestrial) +
  tm_fill(
    col = "lightgreen",
    alpha = .4
  ) +
  tm_shape(ocean) +
  tm_polygons(
    col = "skyblue",
    border.col = "black"
  ) +
  tm_shape(dk2) +
  tm_polygons(
    col = "grey",
    alpha = .8
  ) +
  tm_shape(nature3) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 3
  )

@@

Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about.
A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are variables and indicators.
These are parameters chosen to reflect the central condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 mean that the variable is the worst possible state.
The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthropogenic disturbance.

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021]. This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

@@

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option b in the list above).
Models can describe the general associations between the spatially sampled data and the context (e.g. the set of environmental variables) where it was samples, and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can be very reliable and make good indicators.
This is especially true when the ecosystem assets are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from such models becomes unacceptable.

In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway.
The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.

@@

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variables 1 -5 were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation (ADSV; Figure S1).
This was done by summing the two variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6 is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favorable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

@@

For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
The variable 7GR-GI was rescaled into the indicator named *Trenching* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.

@@

We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
The current study builds on the work in that report.

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference            |
|--------|------------|------------|------------|-----------------|----------------|
| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                             | @runehalvorsen2019   |
| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @runehalvorsen2019   |
| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 6   | 7GR-GI               | Trenching intensity                              | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
| 7   | Infrastructure Index | Infrastructure Index                             | Unitless linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023 |
| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper           |

@@

Because the nature type survey data is spatially biased, we cannot assume that they are area representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 [@tbl-variables, @erikstad_index_2023].
This index is a continuous variable that represnets the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≥ 12.
The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
We plotted the HIA map over an area in Norway that we are familiar with and confirmed a sensible separation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
The resulting distributions are assumed normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^3^ dictated the number of samples for each polygon, thus ensuring that large polygons end up counting more towards the indicator value in the entire municipality.
We then drew 1000 random values from this much large vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e. the ecosystem accounting areas).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

@@

The three indicators showed some association with the HIAs when we lookad at data from all of Norway with `r nrow(naturetypes)` individual mires.
The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
This relationship was much stringer for the indicator Trenching, and weakest for the indicator Alien species.
Alien species also had the highest indicator values overall, with most mirer having no alien species recorded.
For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
However, only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
The credible intervals are widest for the Trenching indicators, and this is because the national data set showed more variation in the underlying variable 7GR-GI, and this information is informing the Bayesian updating process (see @sec-methods).
None of the three municipalities had data for all four HIAs.
Therefore, when we transferred the mean and SD for each HIA over to the ecosystem delineation map (@fig-spread-example), some mire polygons did not get assigned any data, and were therefore dropped from any further analyses.
At the municipality level (i.e. the ecosystem accounting area level) the three municipalities show some differentiation.
Nordre-Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).

@@

```{r fig-EEA-ridgePlot}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Distributions for three ecosystem condition indicators in the Norwegian 
#| municiaplities. The color gradient reflects the value of the x-axis. 
#| The dotted vertical line represnet the threshold for what is considered 
#| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
#| are 2.5%, 50% (the median) and 97.5% percentiles."

>>>>

---
title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

# Survey coverage and other colour overlaid the municipalities
muniPlot <- tm_shape(muni3) +
  tm_borders() +
  tm_facets(
    by = "Municipality",
    ncol = 3
  ) +
  tm_shape(terrestrial) +
  tm_fill(
    col = "lightgreen",
    alpha = .4
  ) +
  tm_shape(ocean) +
  tm_polygons(
    col = "skyblue",
    border.col = "black"
  ) +
  tm_shape(dk2) +
  tm_polygons(
    col = "grey",
    alpha = .8
  ) +
  tm_shape(nature3) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 3
  )

@@

Ecosystem condition accounting is the process of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about.
<!--# About what? -->

A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are *variables* and *indicators*.
These are parameters chosen to reflect the central condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 means that the variable is in the worst possible state.
The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthropogenic disturbance.

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1 <!--# refering to table 1 in Czucz 2021? -->; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021]. This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned a value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

@@

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option b in the list above).
Models can describe the general associations between the spatially sampled data and the context (e.g. the set of environmental variables) where it was sampled, and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can be very reliable and make good indicators.
This is especially true when the ecosystem assets are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from such models becomes unacceptable.

In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighbouring municipalities in Norway.
The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.

@@

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variables 1 -5 were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
This was done by summing the two variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6 is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

@@

For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
The variable 7GR-GI was rescaled into the indicator named *Trenching* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.

@@

We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply due to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recently contributed to a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
The current study builds on the work in that report.

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
|------------|------------|------------|------------|------------|------------|
| 1   | 7FK                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abration                                                              | @runehalvorsen2019   |
| 3   | PRSL                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 4   | 7TK                  | Tracks from large vehicles                       | Unit-less, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks                                                     | @runehalvorsen2019   |
| 5   | PRTK                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 6   | 7GR-GI               | Trenching intensity                              | Unit-less, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
| 7   | Infrastructure Index | Infrastructure Index                             | Unit-less linear scale from 0 to 13.2                                                    | Unit-less index ranging from from 0 to 13.2                                                                                                                       | @erikstad_index_2023 |
| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                    | A categorical representation of the infrastructure index                                                                                                          | this paper           |

@@

Because the nature type survey data is spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 [@tbl-variables, @erikstad_index_2023].
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≥ 12.
The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
We plotted the HIA map over an area in Norway that we are familiar with and confirmed a sensible separation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^3^ dictated the number of samples for each polygon, thus ensuring that large polygons end up counting more towards the indicator value in the entire municipality.
We then drew 1000 random values from this much large vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e. the ecosystem accounting areas).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

@@

The three indicators showed some association with the HIAs when we looked at data from all of Norway with `r nrow(naturetypes)` individual mires.
The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
This relationship was much stronger for the indicator Trenching, and weakest for the indicator Alien species.
Alien species also had the highest indicator values overall, with most mires having no alien species recorded.
For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
However, only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
The credible intervals are widest for the Trenching indicators, and this is because the national data set showed more variation in the underlying variable 7GR-GI, and this information is informing the Bayesian updating process (see @sec-methods).
None of the three municipalities had data for all four HIAs.
Therefore, when we transferred the mean and SD for each HIA over to the ecosystem delineation map (@fig-spread-example), some mire polygons did not get assigned any data, and were therefore dropped from any further analyses.
At the municipality level (i.e. the ecosystem accounting area level) the three municipalities show some differentiation.
Nordre-Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).

@@

```{r fig-EEA-ridgePlot}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Distributions for three ecosystem condition indicators in the Norwegian 
#| municiaplities. The color gradient reflects the value of the x-axis. 
#| The dotted vertical line represent the threshold for what is considered 
#| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
#| are 2.5%, 50% (the median) and 97.5% percentiles."
